---
title: "Quarto Mini Demo: Penguins"
subtitle: "Markdown, R, tables, figures, math, callouts, and diagrams in one short report"
author: "ChatGPT and Jakob Skou Pedersen"
date: 2026-05-07
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
  docx: default
  pdf: default
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(knitr)
library(palmerpenguins)

data <- penguins |> drop_na(species, island, bill_length_mm, flipper_length_mm, body_mass_g)
```

## Why Quarto?

Quarto lets us combine **explanation**, *code*, output, and mathematical notation in one plain-text file. This document uses the `palmerpenguins` data set, which contains `r nrow(data)` complete penguin observations from `r n_distinct(data$species)` species.

::: {.callout-tip}
## Reproducible workflow
Change the code or data, then render again. The numbers, table, and figure update automatically.
:::

We will use a simple linear model idea:

$$
\text{body mass}_i = \beta_0 + \beta_1 \times \text{flipper length}_i + \varepsilon_i
$$

## A small summary table

```{r}
#| label: tbl-summary
#| tbl-cap: "Mean penguin measurements by species."
data |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_bill_mm = mean(bill_length_mm),
    mean_flipper_mm = mean(flipper_length_mm),
    mean_mass_g = mean(body_mass_g),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), round, 1)) |>
  kable()
```

## A figure with a caption

```{r}
#| label: fig-flipper-mass
#| fig-cap: "Larger flippers tend to be associated with higher body mass."
#| fig-alt: "Scatterplot of flipper length versus body mass, colored by penguin species, with one fitted trend line."
ggplot(data, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

As shown in @fig-flipper-mass, the relationship is positive for all three species. You can also refer to @tbl-summary when comparing species averages.

## A tiny workflow diagram

```{mermaid}
flowchart LR
  A[Raw data] --> B[Clean with tidyverse]
  B --> C[Summarise and plot]
  C --> D[Render HTML / PDF / Word]
```

## Try these edits

1. Change `theme: cosmo` to `theme: flatly` in the YAML header.
2. Set `code-fold: false` and render again.
3. Add `pdf: default` under `format:` and try rendering to PDF.
4. Replace `geom_point()` with `geom_jitter()`.
